---
title: "Understanding World Population Dynamics by Huiyu Ding"
subtitle: "Assignment 1 - PSYC593"
format:
  html: default
date: today
author: "Huiyu Ding"
editor: 
  markdown: 
    wrap: 72
output: 
  html_document:
    self_contained: true
    code_folding: hide
    toc: true
---

<!-- These problems come from Imai, K. & Webb Williams, N. (2018). Quantitative Social Science: An introduction -->

<!-- in tidyverse. Princeton University Press. -->

<!-- The code giving partial solutions by Víctor H Cervantes -->

Understanding population dynamics is important for many areas of social
science. We will calculate some basic demographic quantities of births
and deaths for the world's population from two time periods: 1950 to
1955 and 2005 to 2010. We will analyze the following CSV data files -
`Kenya.csv`, `Sweden.csv`, and `World.csv`. Each file contains
population data for Kenya, Sweden, and the world, respectively. The
table below presents the names and descriptions of the variables in each
data set.

<!-- Note the syntax to create a table in markdown -->

<!-- Before the table there must be an empty line, and -->

<!-- the table begins with a line of dashes. -->

<!-- The column names are separated with spaces -->

<!-- and the first character must be a space -->

<!-- Then, segments of repeated dashes indicate the column separations (note -->

<!-- the space between groups of dashes before the character D of the column Description) -->

| Name       | Description                                                                             |
|:----------------------|:------------------------------------------------|
| `country`  | Abbreviated country name                                                                |
| `period`   | Period during which data are collected                                                  |
| `age`      | Age group                                                                               |
| `births`   | Number of births in thousands (i.e., number of children born to women of the age group) |
| `deaths`   | Number of deaths in thousands                                                           |
| `py.men`   | Person-years for men in thousands                                                       |
| `py.women` | Person-years for women in thousands                                                     |

Source: United Nations, Department of Economic and Social Affairs,
Population Division (2013). *World Population Prospects: The 2012
Revision, DVD Edition.*

```{r data_preparation, include = FALSE}
# Load packages ----
library(tidyverse)

# Read data ----
world_data  <- readr::read_csv("./data//raw data/World.csv")
kenya_data  <- readr::read_csv("./data//raw data/Kenya.csv")
sweden_data <- readr::read_csv("./data//raw data/Sweden.csv")
```

<!-- The `read_csv` function from the package `readr` -->

<!-- is part of the `tidyverse`. It will read the data into a -->

<!-- special `data.frame` with more structure known as a `tibble` -->

The data are collected for a period of 5 years where *person-year* is a
measure of the time contribution of each person during the period. For
example, a person that lives through the entire 5 year period
contributes 5 person-years whereas someone who only lives through the
first half of the period contributes 2.5 person-years. Before you begin
this exercise, it would be a good idea to directly inspect each data
set. In R, this can be done with the `View` function, which takes as its
argument the name of a `data.frame` to be examined. Alternatively, in
RStudio, double-clicking a `data.frame` in the `Environment` tab will
enable you to view the data in a spreadsheet-like view.

## Question 1

We begin by computing *crude birth rate* (CBR) for a given period. The
CBR is defined as: $$ 
    \text{CBR} 
     =  \frac{\text{number of births}}{\text{number of person-years lived}}
  $$
<!-- It is also possible to include $\LaTeX$ syntax directly in the document. -->

Compute the CBR for each period, separately for Kenya, Sweden, and the
world. Start by computing the total person-years, recorded as a new
variable within each existing `data.frame` via the `$` operator, by
summing the person-years for men and women. Then, store the results as a
vector of length 2 (CBRs for two periods) for each region with
appropriate labels. You may wish to create your own function for the
purpose of efficient programming. Briefly describe patterns you observe
in the resulting CBRs.

### Answer 1

```{r}
# Create new variable py = total person years for each data set
world_data$py  <- world_data$py.men + world_data$py.women
kenya_data$py  <- kenya_data$py.men + kenya_data$py.women
sweden_data$py <- sweden_data$py.men + sweden_data$py.women
```

<!-- Whenever you find that you will repeat the same calculation -->

<!-- on different sets of data, **consider** creating a function -->

<!-- instead of repeating the code for each piece of data. -->

```{r CBR_function}
# Function to compute the Crude Birth Rate (CBR)
compute_cbr <- function(populationData) {
  populationData %>% 
  group_by(period) %>% 
  summarise(cbr = sum(births) / sum(py)) %>% 
  pull()
}
```

<!-- The "pipe" `%>%` from package `magrittr` takes the output -->

<!-- from the left hand side and makes it the first argument of -->

<!-- the expression on the right hand side. -->

<!-- Starting in R 4.1.0, there is a base pipe operator `|>` -->

<!-- which has a similar behavior. -->

```{r CBR_results}
# Compute the CBR for each data set
(world_cbr  <- compute_cbr(world_data)) # CBR for world
(kenya_cbr  <- compute_cbr(kenya_data)) # CBR for Kenya
(sweden_cbr <- compute_cbr(sweden_data)) # CBR for Sweden
```

There is a general decrease world-wide in terms of crude birth rate.
However, we see opposite trends in Kenya and Sweden, where Kenya has an
increased in crude birth rate while Sweden has a numerical decrease.

## Question 2

The CBR is easy to understand but contains both men and women of all
ages in the denominator. We next calculate the *total fertility rate*
(TFR). Unlike the CBR, the TFR adjusts for age compositions in the
female population. To do this, we need to first calculate the *age
specific fertility rate* (ASFR), which represents the fertility rate for
women of the reproductive age range $[15, 50)$. The ASFR for age range
$[x, x+\delta)$, where $x$ is the starting age and $\delta$ is the width
of the age range (measured in years), is defined as: $$
    \text{ASFR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of births to women of age $[x,\ x+\delta)$}}{\text{Number of person-years lived by women of age $[x,\ x+\delta)$}}
  $$ Note that square brackets, $[$ and $]$, include the limit whereas
parentheses, $($ and $)$, exclude it. For example, $[20, 25)$ represents
the age range that is greater than or equal to 20 years old and less
than 25 years old. In typical demographic data, the age range $\delta$
is set to 5 years. Compute the ASFR for Sweden and Kenya as well as the
entire world for each of the two periods. Store the resulting ASFRs
separately for each region. What does the pattern of these ASFRs say
about reproduction among women in Sweden and Kenya?

### Answer 2

```{r ASFR_function, include = FALSE}
# Function to compute Age specific fertility rate (ASFR)
compute_asfr <- function(pop_data) {
    pop_data %>% 
    mutate(asfr = births / py.women)
}
```

calculate ASFR individually for world, Kenya, and Wweden data.

```{r ASFT_calculation, include = FALSE}
# Compute ASFR for each data set
(world_data  <- compute_asfr(world_data)) # ASFR for world
(kenya_data  <- compute_asfr(kenya_data)) # ASFR for Kenya
(sweden_data <- compute_asfr(sweden_data)) # ASFR for Sweden
```

Compare ASFR between Kenya and Sweden in a graph.

```{r ASDR_Kenya_Sweden}
# Compare ASFRs for Kenya and Sweden and put them in the same df.
(ASFR_Kenya_Sweden_comparison <- data.frame(Kenya = kenya_data$asfr, 
                                            Sweden = sweden_data$asfr))

# make a graph to be able to see the differences visually
ASDR_graph <- ASFR_Kenya_Sweden_comparison %>%
  mutate(age = kenya_data$age,
         period = kenya_data$period) %>%
  pivot_longer(values_to = "ASFR", names_to = "Region", cols = 1:2)

# correct order of ages shown
ASDR_graph$age <- factor(ASDR_graph$age, 
                         levels = gtools::mixedsort(unique(ASDR_graph$age)))

ggplot(data = ASDR_graph, 
       aes(x = age, y = ASFR, group = Region, color = Region)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+ # change the x-ais angles
  ggtitle("ASFR across ages for Sweden, and Kenya in two periods") +
  facet_wrap(~ period) # have two graphs for separate period side-by-side

```

Women in both Kenya and Sweden showed similar pattern that reproduction
starts at 15-19 year-old and decreased across time. Women in Kenya have
higher fertility overall than women in Sweden.

## Question 3

Using the ASFR, we can define the TFR as the average number of children
women give birth to if they live through their entire reproductive age.
$$
  \text{TFR} 
   =   \text{ASFR}_{[15,\ 20)} \times 5 + \text{ASFR}_{[20,\ 25)} \times 5 
  + \dots + \text{ASFR}_{[45,\ 50)} \times 5
  $$

We multiply each age-specific fertility rate rate by 5 because the age
range is 5 years. Compute the TFR for Sweden and Kenya as well as the
entire world for each of the two periods. As in the previous question,
continue to assume that women's reproductive age range is $[15, 50)$.
Store the resulting two TFRs for each country or the world as a vector
of length two. In general, how has the number of women changed in the
world from 1950 to 2000? What about the total number of births in the
world?

### Answer 3

```{r TFR_function, include = FALSE}
# Function to compute the total fertility rate (TFR)
compute_tfr <- function(population_data) {
    population_data %>% 
    group_by(period) %>% 
    summarise(tfr = 5 * sum(asfr)) %>% 
    pull()
}
```

Compute TFR for each data set

```{r TBR_results}
(world_tfr  <- compute_tfr(world_data))
(kenya_tfr  <- compute_tfr(kenya_data))
(sweden_tfr <- compute_tfr(sweden_data))
```

Compare totals of women and births, first calculating them by period and
see how much it has changed.

```{r}
# Compute totals of women and births in the world by period
(totals_world <- world_data %>% 
    group_by(period) %>% 
    summarise(total_women = sum(py.women),
    total_births = sum(births))
)

# Compare how much these totals have changed
(changes_totals <- totals_world[2,-1]/totals_world[1,-1])
```

The total number of women increased from 1950 to 2000, and so does the
total number of births in the world.

## Question 4

Next, we will examine another important demographic process: death.
Compute the *crude death rate* (CDR), which is a concept analogous to
the CBR, for each period and separately for each region. Store the
resulting CDRs for each country and the world as a vector of length two.
The CDR is defined as: $$ 
    \text{CDR} 
     =  \frac{\text{number of deaths}}{\text{number of person-years lived}}
  $$ Briefly describe patterns you observe in the resulting CDRs.

```{r CDR_function}
# Function to compute the Crude death rate (CDR)
compute_cdr <- function(population_data) {
    population_data %>% 
    group_by(period) %>% 
    summarise(cbr = sum(deaths) / sum(py)) %>% 
    pull()
}
```

```{r CDR_for_three_countries}
# Compute the CDR for each data set
(World_cdr  <- compute_cdr(world_data))
(Kenya_cdr  <- compute_cdr(kenya_data))
(Sweden_cdr <- compute_cdr(sweden_data))
```

We are seeing a decreased in crude death rate across the world and
Kenya. However, the crude death rate in Sweden seems to be stable across
time.

## Question 5

One puzzling finding from the previous question is that the CDR for
Kenya during the period of 2005-2010 is about the same level as that for
Sweden. We would expect people in developed countries like Sweden to
have a lower death rate than those in developing countries like Kenya.
While it is simple and easy to understand, the CDR does not take into
account the age composition of a population. We therefore compute the
*age specific death rate* (ASDR). The ASDR for age range $[x, x+\delta)$
is defined as: $$
    \text{ASDR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of deaths for people of age $[x,\ x+\delta)$}}
               {\text{number of person-years of people of age $[x,\ x+\delta)$}}
  $$ Calculate the ASDR for each age group, separately for Kenya and
Sweden, during the period of 2005-2010. Briefly describe the pattern you
observe.

```{r ASDR_function, include = FALSE}
# Function to compute Age specific death rate (ASDR)
compute_asdr <- function(pop_data) {
    pop_data %>% 
    mutate(asdr = deaths/py)
}
```

Calculate ASDR for each region. Then make a graph for ASDR across ages
as a function of region.

```{r ASDR_results}
# Compute ASDR for each data set
world_data_asdr  <- compute_asdr(world_data)
kenya_data_asdr  <- compute_asdr(kenya_data)
sweden_data_asdr <- compute_asdr(sweden_data)

# combine all data into one df to make graph
asdr_results <- data.frame(
    "world_asdr"  = compute_asdr(world_data[world_data$period == "2005-2010",])$asdr,
    "kenya_asdr"  = compute_asdr(kenya_data[kenya_data$period =="2005-2010",])$asdr,
    "sweden_asdr" = compute_asdr(sweden_data[sweden_data$period == "2005-2010",])$asdr,
    "age" = unique(world_data$age)) %>% 
    pivot_longer(names_to = "Region",values_to = "ASDR", 1:3) %>% # long format
    mutate(Region = str_replace_all(Region, "_asdr", "")) # remove '_asdr' 

# correct order of ages shown
asdr_results$age <- factor(asdr_results$age, 
                           levels = gtools::mixedsort(unique(asdr_results$age)))

ggplot(data = asdr_results, aes(x = age, y = ASDR, group = Region, color = Region)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))+
  ggtitle("ASDR across ages for World, Sweden, and Kenya")


```

Sweden across age range has a lower death rate compared to Kenya,
contrary to what we observed when we collapsed across ages.

## Question 6

One way to understand the difference in the CDR between Kenya and Sweden
is to compute the counterfactual CDR for Kenya using Sweden's population
distribution (or vice versa). This can be done by applying the following
alternative formula for the CDR. $$
    \text{CDR}
    \ = \ \text{ASDR}_{[0, 5)} \times P_{[0,5)} + \text{ASDR}_{[5, 10)}
    \times P_{[5, 10)} + \cdots 
  $$ where $P_{[x, x+\delta)}$ is the proportion of the population in
the age range $[x, x+\delta)$. We compute this as the ratio of
person-years in that age range relative to the total person-years across
all age ranges. To conduct this counterfactual analysis, we use
$\text{ASDR}_{[x,x+\delta)}$ from Kenya and $P_{[x,x+\delta)}$ from
Sweden during the period of 2005--2010. That is, first calculate the
age-specific population proportions for Sweden and then use them to
compute the counterfactual CDR for Kenya. How does this counterfactual
CDR compare with the original CDR of Kenya? Briefly interpret the
result.

```{r pop_prop_function}
# Function to compute population proportion by period
compute_pop_prop <- function (pop_data) {
    pop_data %>% 
    group_by(period) %>%
    mutate(pop_prop = py / sum(py)) %>%
    ungroup()
}
```

```{r pop_prop_calculation, include = FALSE}
# Compute population proportion for each data set
(world_data <- compute_pop_prop(world_data))
(kenya_data <- compute_pop_prop(kenya_data))
(sweden_data <- compute_pop_prop(sweden_data))
```

```{r}
# Compute Kenyas CDR Kenya had Sweden's population distribution
kenya_asdr_2005 <- kenya_data_asdr[kenya_data_asdr$period == "2005-2010",]
sweden_data_2005 <- sweden_data[sweden_data$period == "2005-2010",]

kenya_cdr_sweden <- kenya_asdr_2005 %>%
    mutate(temp_cdr = asdr * sweden_data_2005$pop_prop) %>%
    group_by(period) %>%
    summarise(cdr_re_sweden = sum(temp_cdr))
```

The Kenya's CDR to Sweden's population distribution, which gives
`r kenya_cdr_sweden[2]`, is smaller than the `r Kenya_cdr[2]` , which is
the original Kenya's CDR.
